---
title: "Merge Sum Stats"
author: "Sylvia Klosin"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

#========================
# Section 0: Setup
#========================

#=========
# 0.1 Import Packages
#=========
library(data.table)
library(dplyr)
library(here)
library(ggplot2)
library(lubridate)
library(stringr)
library(xtable)

#=========
# 0.2 Set working directory 
#=========
setwd(here::here("RawData"))

#=========
# 0. 3 Importing data
#=========
patent <- fread("invpat_full_disambiguation.csv")
colnames(patent) <- tolower(colnames(patent)) # make all the columns lower case
infutor <- fread("full_DI.csv")

#=========
# 0. 4 Functions
#=========
removeWords <- function(str, stopwords) {
  x <- unlist(strsplit(str, " "))
  paste(x[!x %in% stopwords], collapse = " ")
}

```


# Merge Summary Statistics 

## Filter

In this case we filter the patent and infutor datasets to include only 

- People whose last name starts with ``DI''
- Observations more recent then 2000
- People who live in the United States 


```{r, warning=FALSE}
#========================
# Section 1:  Cleaning patent and infutor data  
#========================
#=========
# 1.1 Making consistent column names
#=========
patent %>% rename(add_city = city,
                  add_state = state, 
                  zip = zipcode, 
                  name_first = firstname,
                  name_last = lastname) -> patent
setDT(patent)

#=========
# 1.2 fix the date variables 
#=========
# for infutor data 
infutor[, addmonth_beg := as.Date(paste(addmonth_beg,"d1",sep=""), format = "%Ym%md%d")]
infutor[, addmonth_end := as.Date(paste(addmonth_end,"d1",sep=""), format = "%Ym%md%d")]
infutor[, first_seen := as.Date(paste(first_seen,"d1",sep=""), format = "%Ym%md%d")]
infutor[, last_seen := as.Date(paste(last_seen,"d1",sep=""), format = "%Ym%md%d")]

# for patent data 
patent[, appdate_clean := as.Date(appdate, "%m/%d/%Y")]
patent[is.na(appdate_clean), appdate_clean := as.Date(appdate, "%Y-%m-%d")] # since the formating of dates is not consistent

#=========
# 1.3 filtering data
#=========
#====
# patent data
#====
# remove people that do not live in the US, we will not have a match for them 
patent <- patent[country == "US",]
# remove observations before 2000
patent <- patent[appyear >= 2000,]
#====
# infutor data
#====
# remove observations before 2000
#infutor[, addmonth_end_year := lubridate::year(addmonth_end)]
#infutor <- infutor[addmonth_end_year >= 2000, ]
```


```{r}
#========================
# Section 2: Making new variables  
#========================
#=======
# 2.1: letter vars based off first and last name
#=======
patent[, last2 := substr(name_last, start = 1, stop = 2)] # first two letters of last name
patent[, first2 := substr(name_first, start = 1, stop = 2)] # first two letters of first name
patent[, first3 := substr(name_first, start = 1, stop = 3)] # first three letters of first name

# cleaning first and middle name
patent[, name_first_clean := gsub("([A-Za-z]+).*", "\\1", name_first)] # remove the weird middle initals
patent[, name_middle_clean := trimws(str_extract(name_first, "\\s\\w"))] # taking the weird middle initals

# cleaning last name 
bad_name_words <- c("JR", "SR", "II", "III", "IV") # words we dont want in last names
patent[, name_last_clean := sapply(name_last,removeWords, stopwords = bad_name_words )] # remove the weird middle initals


infutor[, first2 := substr(name_first, start = 1, stop = 2)] # first two letters of first name
infutor[, first3 := substr(name_first, start = 1, stop = 3)] # first three letters of first name
infutor[, name_first_clean := gsub("([A-Za-z]+).*", "\\1", name_first)] # remove the weird middle initals
infutor[, name_middle_clean := trimws(str_extract(name_first, "\\s\\w"))] # taking the weird middle initals
infutor[, name_last_clean := sapply(name_last,removeWords, stopwords = bad_name_words )]
# since this is the test case # FIXME remove when testing is done
patent <- patent[last2 == "DI",]

#=======
# 2.2: creating new PID vars 
#=======
infutor_base <- infutor[,.(pid, add_id, name_first, name_last, dob)]
infutor_base <- unique(infutor_base[!is.na(dob),])
infutor_base_new <- infutor_base[,pid_new := paste0(pid, collapse = "-"),
                                        .(add_id, name_first, name_last, dob)]
infutor_base_new_ids <- unique(infutor_base_new[,.(pid, pid_new)]) 
# some pid are matched to many pid_new values, where some of the pid_new values are subsets, 
# so we fix this below # do this due to people move, see pid == 135609
infutor_base_new_ids <- unique(infutor_base_new_ids[,pid_new := max(pid_new), .(pid)])
infutor<- merge(infutor, infutor_base_new_ids, by = "pid", all.x = TRUE)
infutor <- infutor[is.na(pid_new), pid_new := as.character(pid)]
```



```{r}
#========================
# Section 3: Removing duplicate information
#========================
# Rows that are identical in the data are not needed in the patent data. 
# Like if a person has multiple patents, but live in the same place for them, 
# we dont need all that data for the merging exersize. 

unique(patent[,.(name_first,
                 name_first_clean,
                 first3,
                 name_last,
                 name_last_clean, 
                 add_city,
                 add_state,
                 unique_inventor_id)]) -> unique_patent
```

## Merge

We merge on last name, state, city and the first three letters of the first name. We have `r length(unique(patent$unique_inventor_id))` unique inventors that we are trying to find matches for. The corresponding infutor set has `r length(unique(infutor$pid))` unique person ids. 

```{r}
#========================
# Section 4: Merging
#========================
patent_infutor_merge <- merge(unique_patent, infutor, 
                              all.x=TRUE, by = c("name_last_clean", "add_city", "add_state","first3"),
                              allow.cartesian=TRUE) # doing a LEFT OUTER JOIN
# returns all the rows from the left table, filling in matched columns (or NA) from the right table
# if there are multiple rows from the right table match to a row in the left table, then new rows 
# will be added to the left. 
```



```{r}
#========================
# Section 5: Merge stats
#========================

patent_infutor_merge_pairs <- unique(patent_infutor_merge[,.(unique_inventor_id, pid_new)])
```


### 1-1 matches 


```{r}
#=========
# How many matches does one inventor get 
#=========
patent_infutor_merge_pairs_no_na <- patent_infutor_merge_pairs[!is.na(pid_new), ]
patent_infutor_merge_pairs_inventor <- patent_infutor_merge_pairs_no_na[,.N,.(unique_inventor_id)]
no_match <- patent[(!(unique_inventor_id %in% patent_infutor_merge_pairs_inventor$unique_inventor_id)),]
# Weird spellings like Allan inventor 04803506-2 % 5
# only first letter in first name is a thing in infutor - is a thing DIGIANFILIPPO inventor 04467844-1 %10 
# werid city 10%
```

We find that doing this merge we get `r sum(!(unique(patent$unique_inventor_id) %in% patent_infutor_merge_pairs_inventor$unique_inventor_id))` inventors that do not get any matches in the infutor data. We were able to lower this number a few different ways. 1) we expanded the infutor time period 2) we cleaned last names by removing things like "JR" and "III".

There are `r sum(patent_infutor_merge_pairs_inventor$N == 1)` inventors that get exactly one match. 

```{r}
patent_one_match <- patent_infutor_merge_pairs_inventor[N == 1,]
patent_one_match <- merge(patent_one_match,
                           patent_infutor_merge,
                           by = "unique_inventor_id")
patent_one_match <- patent_one_match[!is.na(pid_new),]
```




For these 1-1 matches, we get `r sum(patent_one_match$name_first_clean.x == patent_one_match$name_first_clean.y)/nrow(patent_one_match)` have the same clean first name. For those 7% where there is not an exact match, looking at the names show that they seem to be cases where nicknames are used e.g. Fredrick vs Fred.  

### > 1 matches 

For those with more than one match here is a graph that shows the distribution of the number of matches where we truncate the x-axis at 20. There are  `r sum(patent_infutor_merge_pairs_inventor$N > 20)` people with more than 20 matches. 


```{r, results='asis'}
patent_infutor_merge_pairs_inventor %>% filter(N <= 20) %>% ggplot(aes(N)) +
  geom_histogram(binwidth = .5) + 
  theme_bw()
```


```{r, results='asis'}
sum_matches <- data.table(table(patent_infutor_merge_pairs_inventor$N))
colnames(sum_matches) <- c("number_of_mathes", "number_of_people")
sum_matches_table <- xtable(sum_matches, lable = NULL)
print(sum_matches_table, comment=FALSE, include.rownames = FALSE)
```


As a next step we bring in full first name information. 

```{r}
#=========
# People who get some matches
#=========
patent_many_match <- patent_infutor_merge_pairs_inventor[N >= 2,]
patent_many_match <- merge(patent_many_match,
                           patent_infutor_merge,
                           by = "unique_inventor_id")

patent_many_match[, same_first := (name_first_clean.x == name_first_clean.y)]

patent_many_match_first_name_pairs <- patent_many_match[same_first == TRUE,]
patent_many_match_first_name_pairs<- patent_many_match_first_name_pairs[,.N,.(unique_inventor_id)]
```

Adding in the first names only gives us `r sum(patent_many_match_first_name_pairs$N ==1)` more matches.



# ID problems 

We have a problem with PIDS in the infutor data. There are people that are very clearly the same person but have different PID numbers. One such example is Kenneth Dicker from OH. Need to think of a way to reconcile this. For around 300 inventors, it is the case that only the PID seems wrong, and the birthdays and address are exactly the same. It also seems that for many of the rest it the case that PIDs are wrong, and there is a slight typo somewhere.

```{r, results='asis'}
print(xtable(infutor[pid %in% c(304329969, 68862832), .(pid,  dob, add_id, add, add_city, add_state,name_first, name_last 
)]), size="\\tiny", comment = FALSE, include.rownames = FALSE)

```




```{r}
with_pid <- unique(patent_many_match[same_first == TRUE, .(dob, name_first_clean.x, name_first_clean.y,
                                                                              pid_new,add, add_city, name_last_clean, add_state, add_id)])
no_pid <- unique(patent_many_match[same_first == TRUE, .(dob, name_first_clean.x, name_first_clean.y,add, add_city, name_last_clean, add_state, add_id)])

```



```{r}
problems <- unique(with_pid[,.(pid_new, dob, name_first_clean.x, name_last_clean, add_id)])
problems <- with_pid[, new_id := paste0(dob, name_first_clean.x, name_last_clean, add_id)]
problems <- unique(problems[, .(pid_new, new_id)])
sum_problems <- problems[,.N, .(new_id)]
```


If we count the number of people that have the same date of birth, name, and exact address, but different PIDs we end up with `r sum(sum_problems$N > 1)` people.



# Checking Matches 1-1 matches 

```{r}

patent_did_merge <- merge(patent_infutor_merge[pid_new %in% patent_one_match$pid_new,], patent, by = "unique_inventor_id", allow.cartesian = T )
```

## Looking at dates 

```{r}
patent_did_merge[, within_date := ((addmonth_beg - 180) < appdate_clean  & appdate_clean < (addmonth_end + 180))]
```


Seeing for the matches, is there at least one "true" year match. That at least one of the year/addresses in the patent data matches a year/address in the infutor data. We find that this is the case in 1593 cases. The flexibility given is 6 months. s

```{r}

patent_did_merge <- patent_did_merge[, .(sum_match = sum(within_date)), .(unique_inventor_id)]

```



```{r, results='asis'}
#print(xtable(table(patent_did_merge$sum_match != 0)),comment = FALSE)
```


## Looking at middle name

We still need to find the middle name variable in the infutor data. 

```{r}
#patent_did_merge[, same_middle_initial := (name_middle_clean.x == name_middle_clean.y)]

```



# Checking Matches > 1 matches


```{r}

patent_did_merge <- merge(patent[unique_inventor_id %in% patent_many_match$unique_inventor_id,], patent_many_match, by = "unique_inventor_id", allow.cartesian = T )
```

## Looking at dates 

```{r}
patent_did_merge[, within_date := ((addmonth_beg - 180) < appdate_clean  & appdate_clean < (addmonth_end + 180))]
```



```{r}

patent_did_merge_sum <- patent_did_merge[, .(sum_match = sum(within_date)), .(unique_inventor_id, pid_new)]

```


```{r}

patent_did_merge_sum <- patent_did_merge_sum[sum_match != 0 , ]

patent_did_merge_sum  <- patent_did_merge_sum[,(num_match = .N) ,.(unique_inventor_id)]
```

When looking at the many matches, filtering by year only gives us `r sum(patent_did_merge_sum == 1)` more 1 on 1 matches. 









