---
title: "Merge Sum Stats"
author: "Sylvia Klosin"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

#========================
# Section 0: Setup
#========================

#=========
# 0.1 Import Packages
#=========
library(data.table)
library(dplyr)
library(here)
library(ggplot2)
library(lubridate)
library(stringr)
library(xtable)
library(tidyverse)

#=========
# 0.2 Set working directory 
#=========
#setwd(here::here("RawData"))
setwd("~/Desktop")
#=========
# 0. 3 Importing data
#=========
patent <- fread("invpat_full_disambiguation_cityname_cleaned.csv")
#fread("invpat_full_disambiguation.csv")
colnames(patent) <- tolower(colnames(patent)) # make all the columns lower case
infutor <- fread("full_DI_2.csv")
#fread("full_DI.csv")



#=========
# 0. 4 Functions
#=========
removeWords <- function(str, stopwords) {
  x <- unlist(strsplit(str, " "))
  paste(x[!x %in% stopwords], collapse = " ")
}

```


# Merge Summary Statistics 

## Filter

In this case we filter the patent and infutor datasets to include only 

- People whose last name starts with ``DI''
- Observations more recent then 2000 in the patent data
- People who live in the United States 


```{r, warning=FALSE}
#========================
# Section 1:  Cleaning patent and infutor data  
#========================
#=========
# 1.1 Making consistent column names
#=========
patent %>% rename(#add_city = city,
                  #add_state = state, 
                  #zip = zipcode, 
                  name_first = firstname,
                  name_last = lastname, 
                  name_middle = middlename) -> patent
setDT(patent)

#=========
# 1.2 fix the date variables 
#=========
# for infutor data 
infutor[, addmonth_beg := as.Date(paste(addmonth_beg,"d1",sep=""), format = "%Ym%md%d")]
infutor[, addmonth_end := as.Date(paste(addmonth_end,"d1",sep=""), format = "%Ym%md%d")]
infutor[, first_seen := as.Date(paste(first_seen,"d1",sep=""), format = "%Ym%md%d")]
infutor[, last_seen := as.Date(paste(last_seen,"d1",sep=""), format = "%Ym%md%d")]

# for patent data 
patent[, appdate_clean := as.Date(appdate, "%m/%d/%Y")]
patent[is.na(appdate_clean), appdate_clean := as.Date(appdate, "%Y-%m-%d")] # since the formating of dates is not consistent

#=========
# 1.3 filtering data
#=========
#====
# patent data
#====
# remove people that do not live in the US, we will not have a match for them 
patent <- patent[country == "US",]
# remove observations before 2000
patent <- patent[appyear >= 2000,]
#====
# infutor data
#====
# remove observations before 2000
#infutor[, addmonth_end_year := lubridate::year(addmonth_end)]
#infutor <- infutor[addmonth_end_year >= 2000, ]

#=========
# 1.4 fixing the perfered city name
#=========

infutor[pref_city_name == "", pref_city_name := add_city]
patent[pref_city_name == "", pref_city_name := add_city]


```


```{r}
#========================
# Section 2: Making new variables  
#========================
#=======
# 2.1: letter vars based off first and last name
#=======
patent[, last2 := substr(name_last, start = 1, stop = 2)] # first two letters of last name
patent[, first1 := substr(name_first, start = 1, stop = 1)] # first two letters of first name
patent[, first2 := substr(name_first, start = 1, stop = 2)] # first two letters of first name
patent[, first3 := substr(name_first, start = 1, stop = 3)] # first three letters of first name

# cleaning first and middle name
patent %>%
  separate(name_first, c("name_first_clean", "name_middle_clean"), 
           " ", extra = "drop", fill= "right") -> patent

setDT(patent)
patent[, middle1 := substr(name_middle_clean, start = 1, stop = 1)] # first letter middle


# cleaning last name 
bad_name_words <- c("JR", "SR", "II", "III", "IV") # words we dont want in last names
patent[, name_last_clean := sapply(name_last,removeWords, stopwords = bad_name_words )] # remove the weird middle initals

infutor[, first1 := substr(name_first, start = 1, stop = 1)] # first letters of first name
infutor[, first2 := substr(name_first, start = 1, stop = 2)] # first two letters of first name
infutor[, first3 := substr(name_first, start = 1, stop = 3)] # first three letters of first name
infutor[, name_first_clean := name_first] 
infutor[, name_last_clean := name_last]
infutor[, name_middle_clean := name_middle]
infutor[, middle1 := substr(name_middle_clean, start = 1, stop = 1)] # first letter middle

# since this is the test case # FIXME remove when testing is done
patent <- patent[last2 == "DI",]

#=======
# 2.2: creating new PID vars 
#=======
infutor_base <- infutor[,.(pid, add_id, name_first, name_last, dob)]
infutor_base <- unique(infutor_base[!is.na(dob),])
infutor_base_new <- infutor_base[,pid_new := paste0(pid, collapse = "-"),
                                        .(add_id, name_first, name_last, dob)]
infutor_base_new_ids <- unique(infutor_base_new[,.(pid, pid_new)]) 
# some pid are matched to many pid_new values, where some of the pid_new values are subsets, 
# so we fix this below # do this due to people move, see pid == 135609
infutor_base_new_ids <- unique(infutor_base_new_ids[,pid_new := max(pid_new), .(pid)])
infutor<- merge(infutor, infutor_base_new_ids, by = "pid", all.x = TRUE)
infutor <- infutor[is.na(pid_new), pid_new := as.character(pid)]

#=======
# 2.3: creating new inventor ID  
#=======
patent_base <- patent[,.(unique_inventor_id, add_city, add_state, name_first_clean, name_last_clean)]
patent_base_new <- patent_base[,unique_inventor_id_new := paste0(unique_inventor_id, collapse = "-"),
                                        .(add_city, add_state, name_first_clean, name_last_clean)]
patent_base_new_ids <- unique(patent_base_new[,.(unique_inventor_id, unique_inventor_id_new)]) 
patent_base_new_ids <- unique(patent_base_new_ids[,unique_inventor_id_new := 
                                                    max(unique_inventor_id_new), .(unique_inventor_id)])
patent <- merge(patent, patent_base_new_ids, by = "unique_inventor_id", all.x = TRUE)
patent <- patent[is.na(unique_inventor_id_new), unique_inventor_id_new := as.character(unique_inventor_id)]

```



```{r}
#========================
# Section 3: Removing duplicate information
#========================
# Rows that are identical in the data are not needed in the patent data. 
# Like if a person has multiple patents, but live in the same place for them, 
# we dont need all that data for the merging exersize. 

unique(patent[,.(name_first_clean,
                 first3,
                 name_last,
                 name_last_clean, 
                 pref_city_name,
                 add_state,
                 unique_inventor_id_new)]) -> unique_patent
```

## Merge

We merge on last name, state, city and the first three letters of the first name. We have `r length(unique(patent$unique_inventor_id))` unique inventors that we are trying to find matches for. The corresponding infutor set has `r length(unique(infutor$pid))` unique person ids. 

```{r}
#========================
# Section 4: Merging
#========================
patent_infutor_merge <- merge(unique_patent, infutor, 
                              all.x=TRUE, by = c("name_last_clean", "pref_city_name", "add_state","first3"),
                              allow.cartesian=TRUE) # doing a LEFT OUTER JOIN
# returns all the rows from the left table, filling in matched columns (or NA) from the right table
# if there are multiple rows from the right table match to a row in the left table, then new rows 
# will be added to the left. 

```



```{r}
#========================
# Section 5: Merge stats
#========================

patent_infutor_merge_pairs <- unique(patent_infutor_merge[,.(unique_inventor_id_new, pid_new)])
```



### 1-1 matches 


```{r}
#=========
# How many matches does one inventor get 
#=========
patent_infutor_merge_pairs_no_na <- patent_infutor_merge_pairs[!is.na(pid_new), ]
patent_infutor_merge_pairs_inventor <- patent_infutor_merge_pairs_no_na[,.N,.(unique_inventor_id_new)]
no_match <- patent[(!(unique_inventor_id_new %in% patent_infutor_merge_pairs_inventor$unique_inventor_id_new)),]

```

We find that doing this merge we get `r sum(!(unique(patent$unique_inventor_id_new) %in% patent_infutor_merge_pairs_inventor$unique_inventor_id_new))` inventors that do not get any matches in the infutor data. 

(We were able to lower this number a few different ways. 1) we expanded the infutor time period 2) we cleaned last names by removing things like "JR" and "III".)

There are `r sum(patent_infutor_merge_pairs_inventor$N == 1)` inventors that get exactly one match. 

```{r}
patent_one_match <- patent_infutor_merge_pairs_inventor[N == 1,]
patent_one_match <- merge(patent_one_match,
                           patent_infutor_merge,
                           by = "unique_inventor_id_new")
patent_one_match <- patent_one_match[!is.na(pid_new),]
patent_one_match[, same_first := (name_first_clean.x == name_first_clean.y)]
```




For these 1-1 matches, we get `r sum(patent_one_match$name_first_clean.x == patent_one_match$name_first_clean.y)/nrow(patent_one_match)` have the same clean first name. For those 7% where there is not an exact match, looking at the names show that they seem to be cases where nicknames are used e.g. Fredrick vs Fred.  

### > 1 matches 

```{r}
#=========
# People who get some matches
#=========
patent_many_match <- patent_infutor_merge_pairs_inventor[N >= 2,]
patent_many_match <- merge(patent_many_match,
                           patent_infutor_merge,
                           by = "unique_inventor_id_new")

patent_many_match[, same_first := (name_first_clean.x == name_first_clean.y)]

patent_many_match_first_name_pairs <- patent_many_match[same_first == TRUE,
                .(sum_pid_new =length(unique(pid_new))),.(unique_inventor_id_new)]

# bringing in the new matches into "one match"
patent_one_match <- rbind(patent_one_match, patent_many_match[(unique_inventor_id_new %in%
                                           patent_many_match_first_name_pairs[sum_pid_new == 1,
                                           (unique_inventor_id_new)] & (same_first == TRUE)),])
# removing the observations for people that have one right match
patent_many_match <- patent_many_match[!(unique_inventor_id_new %in%
                                           patent_many_match_first_name_pairs[sum_pid_new == 1,
                                           (unique_inventor_id_new)]), ] 


```

#### Full First Name 

As a next step we bring in full first name information. Adding in the first names gives us `r sum(patent_many_match_first_name_pairs$sum_pid_new ==1)` more matches. We have a total of `r length(unique(patent_one_match$unique_inventor_id_new))` 1-1 matches. 


#### Next time we bring in first letter of middle name 

```{r}

patent_did_merge <- merge(patent[unique_inventor_id_new %in% patent_many_match$unique_inventor_id_new,], patent_many_match, by = "unique_inventor_id_new", allow.cartesian = T )
```


```{r}
patent_did_merge[, same_middle := (middle1.x == middle1.y)]

patent_did_merge_sum <- patent_did_merge[, .(sum_match = sum(same_middle)), .(unique_inventor_id_new, pid_new)]

at_least_one_good_middle <- patent_did_merge_sum[sum_match > 0,]
  
one_good_middle <- at_least_one_good_middle[,.N,.(unique_inventor_id_new)][N == 1, .(unique_inventor_id_new)]

# people with only one good date match

one_good_date_pair <- at_least_one_good_middle[unique_inventor_id_new %in% one_good_middle$unique_inventor_id_new,
                                             .(unique_inventor_id_new, pid_new) ]

patent_many_match_good_middle <- merge(one_good_date_pair, patent_many_match, by = c("unique_inventor_id_new", "pid_new"))


# bringing in the new matches into "one match"
patent_one_match <- rbind(patent_one_match, patent_many_match_good_middle)
# removing the observations for people that have one right match
patent_many_match <- patent_many_match[!(unique_inventor_id_new %in% one_good_date_pair$unique_inventor_id_new), ] 

```


Checking dates gives us a total of `r length(unique(patent_one_match$unique_inventor_id_new))` 1-1 matches. Now we have `r length(unique(patent_many_match$unique_inventor_id_new))` greater than 1 matches. 


#### Now we check bad dates

We check if date of application and date of infutor are within plus or minus 180 days. 


```{r}
patent_did_merge[, within_date := ((addmonth_beg - 180) < appdate_clean  & appdate_clean < (addmonth_end + 180))]

patent_did_merge_sum <- patent_did_merge[, .(sum_match = sum(within_date)), .(unique_inventor_id_new, pid_new)]
at_least_one_good_date <-  patent_did_merge_sum[sum_match > 0,]
  
one_good_date <- at_least_one_good_date[,.N,.(unique_inventor_id_new)][N == 1, .(unique_inventor_id_new)]

# people with only one good date match

one_good_date_pair <- at_least_one_good_date[unique_inventor_id_new %in% one_good_date$unique_inventor_id_new,
                                             .(unique_inventor_id_new, pid_new) ]

patent_many_match_good_date <- merge(one_good_date_pair, patent_many_match, by = c("unique_inventor_id_new", "pid_new"))


# bringing in the new matches into "one match"
patent_one_match <- rbind(patent_one_match, patent_many_match_good_date)
# removing the observations for people that have one right match
patent_many_match <- patent_many_match[!(unique_inventor_id_new %in% 
                                           one_good_date_pair$unique_inventor_id_new), ] 

```

Checking dates gives us a total of `r length(unique(patent_one_match$unique_inventor_id_new))` 1-1 matches. Now we have `r length(unique(patent_many_match$unique_inventor_id_new))` greater than 1 matches. 

```{r}

#patent_did_merge_sum <- patent_did_merge_sum[sum_match != 0 , ]
#patent_did_merge_sum  <- patent_did_merge_sum[,(num_match = .N) ,.(unique_inventor_id)]

#When looking at the many matches, filtering by year only gives us `r sum(patent_did_merge_sum == 1)` more 1 on 1 matches. 
```






```{r, results='asis'}

#For those with more than one match here is a graph that shows the distribution of the number of matches where we truncate the #x-axis at 20. There are  `r sum(patent_infutor_merge_pairs_inventor$N > 20)` people with more than 20 matches. 


#sum_matches <- data.table(table(patent_infutor_merge_pairs_inventor$N))
#colnames(sum_matches) <- c("number_of_mathes", "number_of_people")
#sum_matches_table <- xtable(sum_matches, lable = NULL)
#print(sum_matches_table, comment=FALSE, include.rownames = FALSE)
```



```{r, results='asis'}
#patent_infutor_merge_pairs_inventor %>% filter(N <= 20) %>% ggplot(aes(N)) +
#  geom_histogram(binwidth = .5) + 
#  theme_bw()
```



# Checking Matches 1-1 matches 

```{r}

patent_one_one_merge <- merge(patent_infutor_merge[pid_new %in% patent_one_match$pid_new,], patent, by = "unique_inventor_id_new", allow.cartesian = T )
```

## Looking at dates 

```{r}
patent_one_one_merge[, within_date := ((addmonth_beg - 360) < appdate_clean  & appdate_clean < (addmonth_end + 360))]
```

```{r}
patent_did_merge_tab <- patent_one_one_merge[, .(sum_match = sum(within_date)), .(unique_inventor_id_new)]
```


Seeing for the matches, is there at least one "true" year match. That at least one of the year/addresses in the patent data matches a year/address in the infutor data. We find that this is NOT the case for `r sum(patent_did_merge_tab == 0)` inventors. The flexibility given is 1 year. 

```{r}
bad_date_people <- patent_did_merge[unique_inventor_id_new %in%
  patent_did_merge_tab[sum_match == 0,(unique_inventor_id_new)],]
```


```{r, results='asis'}
print(xtable(table(bad_date_people$applyyear)), comment = FALSE)
```

```{r, results='asis'}
#print(xtable(table(patent_did_merge$sum_match != 0)),comment = FALSE)
```




# Second Round Merge to get the people who did not match at all 

Looking at the matches when infutor only has first letter of first name gives us only 5% more matches. 

```{r}
#========================
# Section 4: Merging
#========================
patent_infutor_merge_second <- merge(no_match, infutor, 
                              all.x=TRUE, by = c("name_last_clean", "pref_city_name", "add_state","first1"),
                              allow.cartesian=TRUE) # doing a LEFT OUTER JOIN
# returns all the rows from the left table, filling in matched columns (or NA) from the right table
# if there are multiple rows from the right table match to a row in the left table, then new rows 
# will be added to the left. 

```










